---
title: "core_microbiota"
author: "Ming Yan"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_download: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# Load libraries
```{r}
library(tidyverse)
library(data.table)
```

# Load data
```{r}
# genus level kraken2/bracken results
microbe_genus <- fread("microbial_genus_profile.tsv", header = TRUE) # taxaid is customized, where taxaid > 4294860312 representing taxa that is not represented in the GTDB R207
metadata <- read_csv("metadata_ecology.csv")
metadata_performance <- read_csv("metadata_animal_performance.csv")
metadata_concentrate <- read_csv("metadata_concentrate_level.csv")

# reorder columns
colnames(microbe_genus)[1] <- "Genera"
taxa_rank <- c("Domain", "Phylum", "Class", "Order", "Family", "Genera", "taxaid")
metadata_var <- c("species", "beef_or_dairy", "grazing_or_not", "collection_method", "geography", "project")
sampleid <- colnames(microbe_genus)[!colnames(microbe_genus) %in% taxa_rank]
sample_performane <- metadata_performance$id
new_order <- c(taxa_rank, sampleid)
microbe_genus <- setcolorder(microbe_genus, new_order) # special case for data.table

# recalculate percentage (only bracken results > 0.00001 was kept so total percentage might not sum up 100%)
microbe_genus_abundance <- copy(microbe_genus)
total_percentage <- microbe_genus_abundance %>% select(sampleid) %>% colSums() # average 98%
microbe_genus_abundance[,(sampleid):= lapply(.SD,function(x) x/sum(x)*100), .SDcols = sampleid]

# change to prevelance 
microbe_genus[, (sampleid):=lapply(.SD, as.numeric), .SDcols  = sampleid] # note the parenthesis around sampleid
microbe_genus_prevelance <- microbe_genus[, (sampleid):=lapply(.SD, function(x) ifelse(x>0, 1, 0)), .SDcols = sampleid]
microbe_genus_prevelance$prevelance <- microbe_genus_prevelance %>% select(sampleid) %>% rowSums()
microbe_genus_prevelance$prevelance <- microbe_genus_prevelance$prevelance/976*100
microbe_genus_prevelance <- setcolorder(microbe_genus_prevelance, c(taxa_rank, "prevelance", sampleid))

```

# Load data species level
```{r}
# species level kraken2/bracken results
microbe_species <- fread("microbial_species_profile.tsv", header = TRUE)

# reorder columns
colnames(microbe_species)[1] <- "Species"
taxa_rank_species <- c("Domain", "Phylum", "Class", "Order", "Family", "Genera", "Species", "taxaid")
new_order_species <- c(taxa_rank_species, sampleid)
microbe_species <- setcolorder(microbe_species, new_order_species) # special case for data.table

# recalculate percentage (only bracken results > 0.00001 was kept so total percentage might not sum up 100%)
microbe_species_abundance <- copy(microbe_species)
total_percentage_species <- microbe_species_abundance %>% select(sampleid) %>% colSums()
microbe_species_abundance[,(sampleid):= lapply(.SD,function(x) x/sum(x)*100), .SDcols = sampleid]

# change to prevelance 
microbe_species[, (sampleid):=lapply(.SD, as.numeric), .SDcols  = sampleid] # note the parenthesis around sampleid
microbe_species_prevelance <- microbe_species[, (sampleid):=lapply(.SD, function(x) ifelse(x>0, 1, 0)), .SDcols = sampleid]
microbe_species_prevelance$prevelance <- microbe_species_prevelance %>% select(sampleid) %>% rowSums()
microbe_species_prevelance$prevelance <- microbe_species_prevelance$prevelance/length(sampleid)*100
microbe_species_prevelance <- setcolorder(microbe_species_prevelance, c(taxa_rank_species, "prevelance", sampleid))
```


# get core (100%) and common (50%) taxa 
```{r}
core_genera <- microbe_genus_prevelance[prevelance == 100, Genera]

common_genera <- microbe_genus_prevelance[prevelance > 50, Genera]

core_species <- microbe_species_prevelance[prevelance == 100, Species]

common_species <- microbe_species_prevelance[prevelance > 50, Species]

core_genera_total_abundance <- microbe_genus_abundance[Genera %in% core_genera] %>% as.data.frame() %>% select(sampleid) %>% colSums()

core_species_total_abundance <- microbe_species_abundance[Genera %in% common_species] %>% as.data.frame() %>% select(sampleid) %>% colSums()
```


# core genera abundance, median abundance (log10) > -1 or contain core species
```{r}
core_genus_abundance <- microbe_genus_abundance[Genera %in% core_genera] %>% pivot_longer(cols = sampleid, names_to = "sampleid", values_to = "log10_abundance")
core_genus_abundance$log10_abundance <- log10(core_genus_abundance$log10_abundance)
core_species_abundance <- microbe_species_abundance[Species %in% core_species] %>% pivot_longer(cols = sampleid, names_to = "sampleid", values_to = "log10_abundance")
core_species_abundance$log10_abundance <- log10(core_species_abundance$log10_abundance)

# rank medium abundance
core_genus_abundance_summary <- core_genus_abundance %>% group_by(Genera) %>% summarise(median_log_abundance = median(log10_abundance), max_log_abundance = max(log10_abundance), min_log_abundance = min(log10_abundance)) %>% arrange(median_log_abundance)
abundant_core_genera  <- core_genus_abundance_summary %>% filter(median_log_abundance > -1 | Genera %in% unique(core_species_abundance$Genera)) %>% select(Genera) %>% unlist() # median relative abundance > 0.1%
less_abundant_core_genera  <- core_genus_abundance_summary %>% filter(!Genera %in% abundant_core_genera ) %>% select(Genera) %>% unlist() # median relative abundance > 0.1%
median_core_genera_ordered <- core_genus_abundance_summary$Genera 


# define color palette for both abundant and less abundant core genera. 
phylum_levels <- unique(core_genus_abundance$Phylum)
phylum_palette <- c("#e3af14", "#5ed14f", "#42f5d7", "#0092FF", "#4900FF", "#bf00ff", "#FF00DB", "#ff4000", "#ff8400")
names(phylum_palette) <- phylum_levels

core_genus_abundance_abundant <- core_genus_abundance %>% filter(Genera %in% abundant_core_genera) %>% 
  mutate(Genera = factor(Genera, levels = median_core_genera_ordered), Phylum = factor(Phylum, levels = phylum_levels)) 

core_genus_abundance_abundant$core_species <- ifelse(core_genus_abundance_abundant$Genera %in% unique(core_species_abundance$Genera), "*", "") ## for adding annotation later

core_genus_abundance_abundant %>%  
  ggplot(aes(x=log10_abundance, y=Genera)) + 
  geom_point(aes(colour = Phylum)) + 
  scale_colour_manual(values = phylum_palette, limits = unique(core_genus_abundance %>% filter(Genera %in% abundant_core_genera) %>% select(Phylum) %>% unlist())) + # use limits in scale_colour_manual to display only existed variable, drop = TRUE doesn't work in this version 
  geom_boxplot(outlier.shape = NA, fill = "#f2f6fc") +
  labs(x = "Relative abundance % (log10)", y = "core genera") +
  scale_x_continuous(expand = c(0,0), limits = c(-3.1,2)) +
  expand_limits(y = c(0,43)) + 
  geom_text(aes( x = -3, y = Genera, label = core_species), colour = "red", size = 5, vjust = 0.8) +  # add "*" if core species present in genera; vjust to align "*" with ticks 
  theme(legend.position = "bottom") + 
  theme(legend.title = element_blank()) + 
  guides(color=guide_legend(nrow=3,byrow=TRUE))
            
ggsave("core_genus_abundance_abundant.svg", width = 4.5, height = 7, units = "in")

```



# less abundant core genus
```{r}
core_genus_abundance_less <- core_genus_abundance %>% filter(Genera %in% less_abundant_core_genera) %>% 
  mutate(Genera = factor(Genera, levels = median_core_genera_ordered), Phylum = factor(Phylum, levels = phylum_levels)) 

core_genus_abundance_less$core_species <- ifelse(core_genus_abundance_less$Genera %in% unique(core_species_abundance$Genera), "*", "") ## for adding annotation later

core_genus_abundance_less %>% filter(Genera %in% less_abundant_core_genera) %>% 
  mutate(Genera = factor(Genera, levels = median_core_genera_ordered), Phylum = factor(Phylum, levels = phylum_levels)) %>%  
  ggplot(aes(x=log10_abundance, y=Genera)) + 
  geom_point(aes(colour = Phylum)) + 
  scale_colour_manual(values = phylum_palette, limits = unique(core_genus_abundance %>% filter(Genera %in% less_abundant_core_genera) %>% select(Phylum) %>% unlist())) + # use limits in scale_colour_manual to display only existed variable, drop = TRUE doesn't work in this version 
  geom_boxplot(outlier.shape = NA, fill = "#f2f6fc") +
  labs(x = "Relative abundance % (log10)") +
  scale_x_continuous(expand = c(0.1,0)) + 
  geom_text(aes( x = -3, y = Genera, label = core_species), colour = "red", size = 5) + # add "*" if core species present in genera
  scale_x_continuous(expand = c(0,0), limits = c(-3.1,2)) +
   theme(legend.position = "bottom") + 
  theme(legend.title = element_blank()) + 
  guides(color=guide_legend(nrow=3,byrow=TRUE))

ggsave("core_genus_abundance_rare.svg", width = 4.5, height = 7, units = "in")
```

# core species
```{r}
# rank medium abundance
core_species_abundance_summary <- core_species_abundance %>% group_by(Species) %>% summarise(median_log_abundance = median(log10_abundance), max_log_abundance = max(log10_abundance), min_log_abundance = min(log10_abundance)) %>% arrange(median_log_abundance)
median_core_species_ordered <- core_species_abundance_summary$Species 


# define color palette for both abundant and less abundant core genera. 
genera_levels <- unique(core_species_abundance$Genera)
genera_palette <- c("#e3af14", "#5ed14f", "#42f5d7", "#0092FF", "#4900FF", "#bf00ff")
names(genera_palette) <- genera_levels

core_species_abundance %>%  
  mutate(Species = factor(Species, levels = median_core_species_ordered), Genera = factor(Genera, levels = genera_levels)) %>%  
  ggplot(aes(x=log10_abundance, y=Species)) + 
  geom_point(aes(colour = Genera)) + 
  scale_colour_manual(values = genera_palette, limits = unique(core_species_abundance  %>% select(Genera) %>% unlist())) + # use limits in scale_colour_manual to display only existed variable, drop = TRUE doesn't work in this version 
  geom_boxplot(outlier.shape = NA, fill = "#f2f6fc") +
  labs(x = "Relative abundance % (log10)", y = "core species") +
  theme(legend.position = "bottom") + 
  theme(legend.title = element_blank()) + 
  guides(color=guide_legend(nrow=3,byrow=TRUE))

ggsave("core_species_abundance.svg", width = 4.5, height = 7, units = "in")
```

# plot prevelance of core genera
```{r}
microbe_genus_prevelance %>% ggplot(aes(x = prevelance)) +
  geom_density() +
 geom_histogram(aes(y=after_stat(density)), colour="black", fill="white", bins = 20) +
  theme_minimal()


microbe_genus_prevelance_df <- microbe_genus_prevelance %>% select(-sampleid)
microbe_species_prevelance_df <- microbe_species_prevelance %>% select(-sampleid)
#write_csv(microbe_genus_prevelance_df, "microbe_genus_prevelance_df.csv")
#write_csv(microbe_species_prevelance_df, "microbe_species_prevelance_df.csv")
```

# iterate to check core genera. Are there enough samples to reach plauteu 
```{r}
genus_iterate <- read_csv("core_genus_iteration_res.csv")

colnames(genus_iterate)[1] <- "iteration"
genus_iterate <- genus_iterate %>% pivot_longer(cols = !iteration, names_to = "Sample", values_to = "ncore") %>% mutate(Sample = as.numeric(Sample))

genus_iterate_range <- genus_iterate %>% group_by(Sample) %>% summarise(Min = min(ncore), Max = max(ncore), Mean = mean(ncore)) 
genus_iterate_plot <- left_join(genus_iterate, genus_iterate_range, by="Sample")

genus_iterate_plot %>% ggplot(aes(x=Sample, y=ncore, ymin = Min, ymax = Max, group = iteration)) + 
  geom_ribbon(fill = "grey70", alpha = 0.4) + geom_line(aes(x = Sample, y=Mean), color = "red") +
  scale_x_continuous(breaks = seq(min(genus_iterate_plot$Sample), max(genus_iterate_plot$Sample), by = 100), limits = c(100, 970), expand = c(0,0)) +  # expand to remove extra space outside of the limit
  labs(x = "Number of samples", y = "Number of core genera") +
#theme(axis.text.x = element_text(angle = 90)) + 
 theme_minimal()

ggsave("core_genus_iterration.svg", width = 4.5, height = 3.5, units = "in")
```

# iterate to check core species
```{r}
species_iterate <- read_csv("core_species_iteration_res.csv")

colnames(species_iterate)[1] <- "iteration"
species_iterate <- species_iterate %>% pivot_longer(cols = !iteration, names_to = "Sample", values_to = "ncore") %>% mutate(Sample = as.numeric(Sample))

species_iterate_range <- species_iterate %>% group_by(Sample) %>% summarise(Min = min(ncore), Max = max(ncore), Mean = mean(ncore)) 
species_iterate_plot <- left_join(species_iterate, species_iterate_range, by="Sample")

species_iterate_plot %>% ggplot(aes(x=Sample, y=ncore, ymin = Min, ymax = Max, group = iteration)) + 
  geom_ribbon(fill = "grey70", alpha = 0.4) + geom_line(aes(x = Sample, y=Mean), color = "red") +
  scale_x_continuous(breaks = seq(min(genus_iterate_plot$Sample), max(genus_iterate_plot$Sample), by = 100), limits = c(100, 970), expand = c(0,0)) +  # expand to remove extra space outside of the limit
  labs(x = "Number of samples", y = "Number of core species") + 
  #theme(axis.text.x = element_text(angle = 90)) + 
  theme_minimal()

ggsave("core_species_iterration.svg", width = 4.5, height = 3.5, units = "in")
```

# prepare data for DAA, only core
```{r}
core_species <- core_species_abundance %>% filter(sampleid %in% sample_performane) %>% mutate(abundance = 10^log10_abundance/100) %>% 
  select(-log10_abundance) %>% select(-c(Domain, Phylum, Class, Order, Family, Genera, taxaid)) %>% pivot_wider(names_from = Species, values_from = abundance)

core_genus_abundancea_performance <- core_genus_abundance %>% filter(sampleid %in% sample_performane) %>% mutate(abundance = 10^log10_abundance/100) %>% select(-log10_abundance) %>% 
   select(-c(Domain, Phylum, Class, Order, Family, taxaid)) %>% pivot_wider(names_from = Genera, values_from = abundance)

core_species_taxa <- core_species_abundance %>% select(Domain, Phylum, Class, Order, Family, Genera, Species) %>% unique()

core_genus_taxa <- core_genus_abundance %>% select(Domain, Phylum, Class, Order, Family, Genera) %>% unique()

write_csv(core_genus_abundancea_performance, "core_genus_abundance_animal_performance.csv")
write_csv(core_species, "core_species_abundance_animal_performance.csv")
write_csv(core_species_taxa, "core_species_taxa.csv")
write_csv(core_genus_taxa, "core_genus_taxa.csv")
```

# prepare data for DAA, all
```{r}
sample_performane <- intersect(sample_performane, sampleid)
sample_performane_genus_columns <- c("Genera", sample_performane)
sample_performane_species_columns <- c("Species", sample_performane)

# get taxa
genus_taxa <- microbe_genus_abundance %>% select(Domain, Phylum, Class, Order, Family, Genera,taxaid) %>% unique()
species_taxa <- microbe_species_abundance %>% select(Domain, Phylum, Class, Order, Family, Genera, Species, taxaid) %>% unique()

# change to relative abundance and transpose (for Maaslin)
microbe_genus_abundance_df <- microbe_genus_abundance[, (sampleid) := .SD/100, .SDcols = sampleid] %>% as.data.frame()
microbe_genus_abundance_df_performance <- microbe_genus_abundance_df %>% select(sample_performane_genus_columns)
write_csv(microbe_genus_abundance_df_performance, "full_genus_abundance_animal_performance_linda.csv") # for linda


rownames(microbe_genus_abundance_df_performance) <- microbe_genus_abundance_df_performance[,1]
microbe_genus_abundance_df_performance[,1] <- NULL 
rownames_t <- colnames(microbe_genus_abundance_df_performance)
colnames_t <- rownames(microbe_genus_abundance_df_performance)
microbe_genus_abundance_df_performance_t <- transpose(microbe_genus_abundance_df_performance)
rownames(microbe_genus_abundance_df_performance_t) <- rownames_t
colnames(microbe_genus_abundance_df_performance_t) <- colnames_t

# repeat for species level
microbe_species_abundance_df <- microbe_species_abundance[, (sampleid) := .SD/100, .SDcols = sampleid] %>% as.data.frame()
microbe_species_abundance_df_performance <- microbe_species_abundance_df %>% select(sample_performane_species_columns)
write_csv(microbe_species_abundance_df_performance, "full_species_abundance_animal_performance_linda.csv") # for linda
rownames(microbe_species_abundance_df_performance) <- microbe_species_abundance_df_performance[,1]
microbe_species_abundance_df_performance[,1] <- NULL 
rownames_t <- colnames(microbe_species_abundance_df_performance)
colnames_t <- rownames(microbe_species_abundance_df_performance)
microbe_species_abundance_df_performance_t <- transpose(microbe_species_abundance_df_performance)
rownames(microbe_species_abundance_df_performance_t) <- rownames_t
colnames(microbe_species_abundance_df_performance_t) <- colnames_t

# write to csv
#write_csv(microbe_genus_abundance_df_performance_t, "full_genus_abundance_animal_performance.csv")
#write_csv(microbe_species_abundance_df_performance_t, "full_species_abundance_animal_performance.csv")

#write_csv(species_taxa, "species_taxa.csv")
#write_csv(genus_taxa, "genus_taxa.csv")

```


# session info
```{r}
sessionInfo()
```








