---
title: "microbe_SpiecEasi"
author: "Ming Yan"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_download: true
---


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# Load libraries
```{r}
library(tidyverse)
library(SpiecEasi)
library(igraph)
library(stringr)
library(RColorBrewer)
```

## R Markdown
```{r}
library(tidyverse)
library(SpiecEasi)

microbe_50 = read_csv("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/microbes_prevalence_filtered_50.csv")
virome_50 = read_csv("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/virome_prevalence_filtered_50.csv")
microbe_taxa = read_csv("/fs/ess/PAS0439/MING/databases/hungate_rug_rumen_mags_taxa.csv")
core_genus_df = read_csv("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/core_genus_abundant.csv")
core_species_df = read_csv("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/core_species_taxa.csv")
rvd_host = read_tsv("/fs/ess/PAS0439/MING/virome/checkv_trimmed_for_dowmstream/RVD_resources/host_match.tsv")
rvd_host$species = gsub("s__", "", rvd_host$species)
se.micro_50 = readRDS("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/se.micro_50.rds")
se.rvd_50 = readRDS("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/se.rvd_50.rds")
core_genus = core_genus_df %>% select(Genera) %>% unlist()

taxa = read_csv("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/taxa.csv")

```

```{r}
# color nodes based on phylum
find_phylum <- function(x) {
   Phylum = taxa %>%  filter(Species == x) %>% select(Phylum) %>% unique() %>% unlist()
   
   if(length(Phylum) == 0 ){
     return("Unknown")
   }
   else if(Phylum == "Methanobacteriota"){
      return("Archaea")
   }else{
     return(Phylum)
   }
   
}

find_host_phylum <- function(x) {
   Phylum = rvd_host %>%  filter(sseqid == x) %>% select(phylum) %>% unique() %>% unlist()
   
   if(length(Phylum) == 0 ){
     return("Unknown")
   }
   else if(str_split(Phylum, "__")[[1]][2] == "Methanobacteriota"){
      return("Archaea")
   } else{
     return(str_split(Phylum, "__")[[1]][2])
   }
   
}

```


## preprocessing
```{r}

microbe_phylum_50 <-  sapply(colnames(microbe_50)[2:ncol(microbe_50)], find_phylum) 
micro_phylum = unique(microbe_phylum_50)
rvd_host_phylum <- sapply(colnames(virome_50)[2:ncol(virome_50)], find_host_phylum)  

mycolors_50 = c("#4DAF4A", "#984EA3", "#FF7F00", "grey", "#1F78B4", "#B2DF8A", "black", "#FB9A99", "#E31A1C",
                "#FDBF6F", "#8DD3C7", "#FFFFB3", "#1c8a94", "#FB8072", "#80B1D3", "#FDB462", "#fcba03")
names(mycolors_50) = c("Firmicutes_C", "Bacteroidota", "Firmicutes_A", "Unknown", "Spirochaetota",     
 "Verrucomicrobiota", "Archaea", "Proteobacteria", "Actinobacteriota", "Chloroflexota",     
 "Firmicutes", "Fibrobacterota", "Cyanobacteria", "Planctomycetota", "Desulfobacterota_I", "Thermoplasmatota", "Synergistota")
 
phylum = c(rvd_host_phylum, microbe_phylum_50)


for (f in 1:length(mycolors_50)){
     phy = names(mycolors_50)[f]
     col = mycolors_50[f]
   
   phylum = sapply(phylum, function(x){
      gsub(phy, col, x)
   })
   }
    
core_species = intersect(core_species_df$Species, colnames(microbe_50))

microbe_node_size_50 = c()
for (f in colnames(microbe_50)[2:ncol(microbe_50)]){
   if (f %in% core_species){
      microbe_node_size_50 = c(microbe_node_size_50,6)
   }else {
      microbe_node_size_50 = c(microbe_node_size_50,4)
   }
}



```

## network microbe only 
```{r}

igraph.micro_50 = adj2igraph(getRefit(se.micro_50))
igraph.rvd_50 = adj2igraph(getRefit(se.rvd_50))

V(igraph.micro_50)$phylum_color = microbe_phylum_50


mn_modularity = fastgreedy.community(igraph.micro_50)

mn_table = table(mn_modularity$membership)
module_kept_mn = names(mn_table[mn_table > rev(sort(mn_table))[4]]) # kept the largest 3 modules 
vertices_kept_mn = which(mn_modularity$membership %in% module_kept_mn)

mn_membership_v = mn_modularity$membership
names(mn_membership_v) = mn_modularity$names

module_kept_list_mn = list()
for (f in 1:length(module_kept_mn)){
   module = module_kept_mn[f]
   module_kept_list_mn[[f]] = names(mn_membership_v[mn_membership_v == module])
}

E(igraph.micro_50)$color = "blue"

set.seed(1)
svg(filename="/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/microbial_network_50_prevalence.svg", width = 3.5, height = 3.5)
par(mar = c(0.1, 0.1, 0.1, 0.1))
plot(igraph.micro_50, vertex.size=microbe_node_size_50, vertex.label = NA,
     vertex.color=phylum[(length(rvd_host_phylum)+1):length(phylum)],
     mark.groups=module_kept_list_mn, 
     mark.col = rainbow(length(module_kept_list_mn), alpha = 0.3), 
     mark.expand=0,
     mark.border=NA
     )
dev.off() 
```

## microbial network modularity and assortativity 
```{r}
V(igraph.micro_50)$phylum_as_numeric = as.numeric(factor(phylum[(length(rvd_host_phylum)+1):length(phylum)]))
assortativity(igraph.micro_50, V(igraph.micro_50)$phylum_as_numeric, directed=F)

mn_modularity

```

## def attribute for cross-domain network
```{r}
Type = c(rep("Virus", ncol(virome_50) - 1), rep("Microbe", ncol(microbe_50) - 1))

V(igraph.rvd_50)$Type = Type
V(igraph.rvd_50)$phylum_color = phylum


E(igraph.rvd_50)[1:570 %--% 1:570 ]$type = "vtov"
E(igraph.rvd_50)[571:1241 %--% 571:1241 ]$type = "mtom"
E(igraph.rvd_50)[1:571 %--% 571:1241 ]$type = "vtom"

edge_color = c()
for (f in E(igraph.rvd_50)$type){
   if (f == "vtov"){
      edge_color = c(edge_color, "green")
   } else if (f == "vtom"){
      edge_color = c(edge_color, "red")
   } else {
     edge_color = c(edge_color, "blue") 
   }
}

E(igraph.rvd_50)$color = edge_color

microbe_node_size = c()
for (f in colnames(microbe_50)[2:ncol(microbe_50)]){
   if (f %in% core_species){
      microbe_node_size = c(microbe_node_size,6)
   }else {
      microbe_node_size = c(microbe_node_size,4)
   }
}

core_species_rvd = rvd_host %>% filter(species %in% core_species) %>% select(sseqid)

virome_node_size = c()
for (f in colnames(virome_50)[2:ncol(virome_50)]){
   if (f %in% core_species_rvd){
      virome_node_size = c(virome_node_size,6)
   }else {
      virome_node_size = c(virome_node_size,4)
   }
}

vertex_size_v = c(virome_node_size, microbe_node_size)


```

## add module info 
```{r}
cdn_modularity = fastgreedy.community(igraph.rvd_50)

cdn_table = table(cdn_modularity$membership)
module_kept = names(cdn_table[cdn_table > rev(sort(cdn_table))[4]]) # kept the largest 3 modules  
vertices_kept = which(cdn_modularity$membership %in% module_kept)

cdn_membership_v = cdn_modularity$membership
names(cdn_membership_v) = cdn_modularity$names

module_kept_list = list()
for (f in 1:length(module_kept)){
   module = module_kept[f]
   module_kept_list[[f]] = names(cdn_membership_v[cdn_membership_v == module])
}


set.seed(1)
svg(filename="/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/cross_domain_network_50_prevalence.svg", width = 3.5, height = 3.5)
par(mar = c(0.1, 0.1, 0.1, 0.1))

plot(igraph.rvd_50, vertex.size=vertex_size_v, vertex.label = NA, 
     vertex.shape=ifelse(V(igraph.rvd_50)$Type == "Virus", "square", "circle"),
     vertex.color=phylum,
     mark.groups=module_kept_list, 
     mark.col = c("#0000FF4D","#FF00004D", "#00FF004D"),
     mark.border=NA,
      mark.expand=0
     )
dev.off() 
```

```{r}
V(igraph.rvd_50)$phylum_as_numeric = as.numeric(factor(phylum))
assortativity(igraph.rvd_50, V(igraph.rvd_50)$phylum_as_numeric, directed=F)

cdn_modularity
```

## get average degree between any two bacterial nodes, cross-domain network vs. microbial only network 
```{r}

cdn_degree = read_csv("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/cdn_degree_50.csv")
mn_degree = data_frame(node = names(degree(igraph.micro_50)), degree = degree(igraph.micro_50))

df = merge(cdn_degree, mn_degree, by = "node")

colnames(df) = c("node", "cross_domain", "microbial")
df_ready = df %>% pivot_longer(c(cross_domain, microbial), names_to = "type", values_to = "degree")


t.test(df$cross_domain, df$microbial, paired = TRUE, alternative = "greater")

mean(df$cross_domain)
sd(df$cross_domain)

mean(df$microbial)
sd(df$microbial)
```

## number of singleton 
```{r}
df %>% filter(cross_domain == 0) %>% nrow()
df %>% filter(microbial == 0) %>% nrow()
```

## keystone species
```{r}
bet = betweenness(igraph.rvd_50)
deg = degree(igraph.rvd_50)
microbe_species = colnames(microbe_50)[2:ncol(microbe_50)]
votus = colnames(virome_50)[2:ncol(virome_50)]
species = c(votus, microbe_species)
type = c(rep("Virus", length(votus)), rep("Microbe", length(microbe_species)))

keystone = data.frame(betweenness = bet, degree = deg, species = species, type = type, phylum = phylum)

keystone %>% ggplot(aes(degree, betweenness)) + geom_point(aes(color = phylum)) + 
  scale_color_identity(guide = "legend") + facet_wrap(~ type) +
   theme_bw() +
   theme(legend.position =  "none") +
   theme(text = element_text(size=10),
   axis.text.x = element_text(size=8),
   axis.text.y = element_text(size=8),
   axis.title.x = element_text(size=8),
   axis.title.y = element_text(size=8)
) 
  
ggsave("/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/keystone_species.svg", width = 7, height = 4)

```

## cross-domain network too crowded, seperated to see details.
```{r}
set.seed(1)
## 3,4,6
g = igraph.rvd_50
c =  fastgreedy.community(igraph.rvd_50)
l <- layout_nicely(g)

module1_id = 3
module2_id = 4
module3_id = 6

module1_v_idxs <- which(c$membership == module1_id)
module2_v_idxs <- which(c$membership == module2_id)
module3_v_idxs <- which(c$membership == module3_id)

g_sub_module1 <- induced_subgraph(g, V(g)[module1_v_idxs])
g_sub_module2 <- induced_subgraph(g, V(g)[module2_v_idxs])
g_sub_module3 <- induced_subgraph(g, V(g)[module3_v_idxs])


module1_edge_col = c()
for (f in E(g_sub_module1)$type){
   if (f == "vtov"){
      module1_edge_col = c(module1_edge_col, "green")
   } else if (f == "vtom"){
      module1_edge_col = c(module1_edge_col, "red")
   } else {
     module1_edge_col = c(module1_edge_col, "blue") 
   }
}


module2_edge_col = c()
for (f in E(g_sub_module2)$type){
   if (f == "vtov"){
      module2_edge_col = c(module2_edge_col, "green")
   } else if (f == "vtom"){
      module2_edge_col = c(module2_edge_col, "red")
   } else {
     module2_edge_col = c(module2_edge_col, "blue") 
   }
}

module3_edge_col = c()
for (f in E(g_sub_module3)$type){
   if (f == "vtov"){
      module3_edge_col = c(module3_edge_col, "green")
   } else if (f == "vtom"){
      module3_edge_col = c(module3_edge_col, "red")
   } else {
     module3_edge_col = c(module3_edge_col, "blue") 
   }
}

set.seed(1)
svg(filename="/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/cross_domain_network_50_prevalence_module1.svg", width = 2.5, height = 2.5, bg = "transparent")
par(mar = c(0.1, 0.1, 0.1, 0.1))
plot(g_sub_module1,
  layout = l[module1_v_idxs, ],
  edge.color = module1_edge_col,
  mark.border = NA,
  mark.col = rainbow(3, alpha = 0.3)[2],
  mark.groups = V(g_sub_module1),
  vertex.label = NA,
  vertex.color=phylum[module1_v_idxs],
  vertex.size=vertex_size_v[module1_v_idxs], vertex.label = NA, 
   vertex.shape=ifelse(V(g_sub_module1)$Type == "Virus", "square", "circle")
)
dev.off()


set.seed(1)
svg(filename="/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/cross_domain_network_50_prevalence_module2.svg", width = 2.5, height = 2.5, bg = "transparent")
par(mar = c(0.1, 0.1, 0.1, 0.1))
plot(g_sub_module2,
  layout = l[module2_v_idxs, ],
  edge.color = module2_edge_col,
  mark.border = NA,
  mark.col = rainbow(3, alpha = 0.3)[1],
  mark.groups = V(g_sub_module2),
  vertex.label = NA,
  vertex.color=phylum[module2_v_idxs],
  vertex.size=vertex_size_v[module2_v_idxs], vertex.label = NA, 
   vertex.shape=ifelse(V(g_sub_module2)$Type == "Virus", "square", "circle")
)
dev.off()

set.seed(1)
svg(filename="/fs/ess/PAS0439/MING/virome_ecology/results/R_project/virome_microbes_network_RUG4941/cross_domain_network_50_prevalence_module3.svg", width = 2.5, height = 2.5, bg = "transparent")
par(mar = c(0.1, 0.1, 0.1, 0.1))
module3_plot = plot(g_sub_module3,
  layout = l[module3_v_idxs, ],
  edge.color = module3_edge_col,
  mark.border = NA,
  mark.col = rainbow(3, alpha = 0.3)[3],
  mark.groups = V(g_sub_module3),
  vertex.label = NA,
  vertex.color=phylum[module3_v_idxs],
  vertex.size=vertex_size_v[module3_v_idxs], vertex.label = NA, 
   vertex.shape=ifelse(V(g_sub_module3)$Type == "Virus", "square", "circle")
)
dev.off()


```





